---
title: "Assignment 2 - Report"
author: "Eleni Liarou, Zoë Azra Blei, Frederieke Loth, group 20"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
fontsize: 11pt
highlight: tango
---

::: {style="text-align: justify;"}
# Exercise 1: Titanic

#### Section a

#### Section b

#### Section c

#### Section d

#### Section e

# Exercise 2: Military Coups

```{r echo=FALSE}
library(MASS)
data <- read.table("coups.txt", header = TRUE, row.names = 1)

# Transform 'pollib' into a factor, as we think it may not have a linear effect
data$pollib <- as.factor(data$pollib)
```

#### Section a
Note that we transformed 'pollib' into a factor since we predicted the effect to be non-linear. This was found to be correct after comparing the different Poisson model outcomes. 'pollib = 1' and 'pollib = 2' are henceforth compared to the value of 'pollib = 0'

``` {r echo=FALSE}
poisson_reg <- glm (miltcoup ~ ., data = data, family = poisson)
summary(poisson_reg)
```

We determine that variables for which p-value > 0.05 do not have a statistically significant effect on our response variable 'miltcoup' which signifies the number of military coups. Moreover, a positive coefficient details that an increase in the corresponding variable indicates an increase in 'miltcoup'. We can therefore state the following: 'oligarchy' and 'parties' are statistically significant with a positive effect on 'miltcoup'. 'pollib1' has a slightly significant and negative effect on 'miltcoup' whilst 'pollib2' has a significant negative effect. We can thus say that according to the data, countries with a higher political liberation factor may experience less military coups.

# Remove insignificant pieces one by one (summary)
#### Section b

#### Section c

# Exercise 3: Stormer viscometer

#### Section a

#### Section b

#### Section c

#### Section d

#### Section e
:::
